---
title: "Machine Learning"
author: "Elad Deiss-Yehiely"
date: "2026-04-01"
output: html_document
---




######### EDY 4/1/26 HERE FOR TRYING TO IMPLEMENT SLIGHT FIXES OF FEATURE SELECTION FROM CLAUDE #########
```{r}
# Overview

#This document trains a suite of machine learning classifiers to predict
#chronic_rhinitis24m_current (binary: 0/1) using ASV relative abundances
#as predictors. Models are evaluated by repeated k-fold cross-validation;
#primary metric is **ROC AUC**.


# 1. Libraries & Configuration
library(tidyverse)
library(phyloseq)
library(caret)
library(glmnet)
library(randomForest)
library(pls)
library(earth)
library(gbm)
library(ipred)
library(pROC)       # for plotting ROC curves
library(knitr)
library(kableExtra)

# ── USER CONFIGURATION ────────────────────────────────────────────────────────
PHYLOSEQ_PATH  <- "/Users/ediessy/Desktop/phy_prev_10O_10R.rds"       # <-- update this path
OUTCOME_COL    <- "chronic_rhinitis36m_current"
SUBJECT_ID_COL <- "Subject_ID"
POSITIVE_CLASS <- "Yes"                      # the level treated as the event
                                             # change to "1" if stored as 0/1
SEED           <- 481516
RESULTS_DIR    <- "/Users/ediessy/Desktop/Postdoc UCSF/CANOE/results_asv_ml"
# ─────────────────────────────────────────────────────────────────────────────

dir.create(RESULTS_DIR, showWarnings = FALSE, recursive = TRUE)
```

# 2. Load & Prepare Data

## 2.1 Load phyloseq and convert to relative abundance

```{r load-data}
ps <- readRDS(PHYLOSEQ_PATH)
cat("Phyloseq object loaded:\n")
print(ps)

# Convert to relative abundance
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
cat("\nConverted to relative abundance.\n")
```

## 2.2 Extract ASV feature matrix

```{r feature-matrix}
# OTU table: samples as rows
asv_mat <- as.data.frame(otu_table(ps_rel))
if (!taxa_are_rows(ps_rel)) {
  # already samples × ASVs — nothing to do
} else {
  asv_mat <- as.data.frame(t(asv_mat))
}

# Clean column names so caret doesn't choke on special characters
colnames(asv_mat) <- make.names(colnames(asv_mat), unique = TRUE)

cat(sprintf("ASV matrix: %d samples × %d ASVs\n", nrow(asv_mat), ncol(asv_mat)))
```

## 2.3 Extract outcome

```{r extract-outcome}
meta <- as.data.frame(sample_data(ps_rel))

if (!OUTCOME_COL %in% colnames(meta)) {
  stop(sprintf(
    "Outcome column '%s' not found in sample_data. Available columns:\n%s",
    OUTCOME_COL, paste(colnames(meta), collapse = ", ")
  ))
}

outcome_raw <- as.character(meta[[OUTCOME_COL]])

# Recode TRUE/FALSE → valid R names
outcome_raw <- dplyr::recode(outcome_raw, "TRUE" = "Yes", "FALSE" = "No")

# Coerce to factor with two levels; POSITIVE_CLASS will be the second level
# so caret's twoClassSummary treats it as the event of interest
outcome_factor <- factor(
  outcome_raw,
  levels = c("No", "Yes")
)
cat(sprintf("Outcome levels: %s  (positive class = '%s')\n",
            paste(levels(outcome_factor), collapse = " / "), POSITIVE_CLASS))
cat("Class distribution:\n")
print(table(outcome_factor, useNA = "always"))
```

## 2.4 Align samples and create final X / y

```{r align-samples}
# Use only samples present in both the ASV matrix and the metadata
shared_samples <- intersect(rownames(asv_mat), rownames(meta))
cat(sprintf("Samples in ASV matrix: %d\n", nrow(asv_mat)))
cat(sprintf("Samples in metadata:   %d\n", nrow(meta)))
cat(sprintf("Shared samples:        %d\n", length(shared_samples)))

X <- asv_mat[shared_samples, , drop = FALSE]
y <- outcome_factor[match(shared_samples, rownames(meta))]

# Drop samples with missing outcome
complete_idx <- !is.na(y)
X <- X[complete_idx, , drop = FALSE]
y <- droplevels(y[complete_idx])

cat(sprintf("\nFinal dataset after removing NA outcomes: %d samples × %d ASVs\n",
            nrow(X), ncol(X)))
cat("Final class distribution:\n")
print(table(y))

Xm <- as.matrix(X)   # numeric matrix for glmnet
```

---

# 3. Cross-Validation Setup

```{r train-control}
n <- nrow(X)
p <- ncol(X)

# k: at least 5 samples per fold; cap at 10
k       <- min(10, max(3, floor(n / 5)))
repeats <- 5

cat(sprintf("Cross-validation: %d-fold, %d repeats  (n = %d)\n", k, repeats, n))

ctrl <- trainControl(
  method           = "repeatedcv",
  number           = k,
  repeats          = repeats,
  classProbs       = TRUE,           # required for ROC
  summaryFunction  = twoClassSummary,
  savePredictions  = "final",
  returnResamp     = "all",
  allowParallel    = TRUE,
  verboseIter      = FALSE
)

# Primary optimisation metric
METRIC <- "ROC"
```

> **Note on class imbalance:** If classes are severely imbalanced (< 20 % minority),
> consider adding `sampling = "up"` (upsampling) or `sampling = "smote"` to
> `trainControl()` above, and install the **DMwR2** package for SMOTE.

---

# 4. Train Models

```{r train-models, results='hold'}
set.seed(SEED)
models     <- list()
lambda_grid <- 10^seq(-4, 0, length.out = 20)

safe_train <- function(label, expr) {
  cat(sprintf("Training %s ...\n", label))
  tryCatch({
    m <- eval(expr)
    best_roc <- max(m$results[[METRIC]], na.rm = TRUE)
    cat(sprintf("  OK  best ROC AUC: %.4f\n", best_roc))
    m
  }, error = function(e) {
    cat(sprintf("  FAIL: %s\n", e$message))
    NULL
  })
}

# ── 1. Ridge (alpha = 0, logistic via glmnet) ──────────────────────────────
models$ridge <- safe_train("Ridge (glmnet, alpha=0)", quote(
  train(x = Xm, y = y, method = "glmnet", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(alpha = 0, lambda = lambda_grid),
        preProcess = c("center", "scale"), family = "binomial")
))

# ── 2. Lasso (alpha = 1) ───────────────────────────────────────────────────
models$lasso <- safe_train("Lasso (glmnet, alpha=1)", quote(
  train(x = Xm, y = y, method = "glmnet", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(alpha = 1, lambda = lambda_grid),
        preProcess = c("center", "scale"), family = "binomial")
))

# ── 3. Elastic Net (alpha tuned) ───────────────────────────────────────────
models$elastic_net <- safe_train("Elastic Net (glmnet, alpha tuned)", quote(
  train(x = Xm, y = y, method = "glmnet", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(alpha = c(0.25, 0.5, 0.75), lambda = lambda_grid),
        preProcess = c("center", "scale"), family = "binomial")
))

# ── 4. PLS-DA ──────────────────────────────────────────────────────────────
max_comp <- min(15, p, n - 2)
models$plsda <- safe_train("PLS-DA", quote(
  train(x = Xm, y = y, method = "pls", metric = METRIC, trControl = ctrl,
        tuneGrid   = data.frame(ncomp = 1:max_comp),
        preProcess = c("center", "scale"))
))

# ── 5. MARS ────────────────────────────────────────────────────────────────
models$mars <- safe_train("MARS (earth)", quote(
  train(x = Xm, y = y, method = "earth", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(degree = 1:2, nprune = c(5, 10, 15, 20)),
        preProcess = c("center", "scale"),
        glm = list(family = binomial))   # logistic MARS
))

# ── 6. Random Forest ───────────────────────────────────────────────────────
mtry_grid <- unique(c(2, floor(sqrt(p)), floor(p / 3), floor(p / 2)))
mtry_grid <- sort(mtry_grid[mtry_grid >= 1 & mtry_grid <= p])
models$random_forest <- safe_train("Random Forest", quote(
  train(x = Xm, y = y, method = "rf", metric = METRIC, trControl = ctrl,
        tuneGrid   = data.frame(mtry = mtry_grid),
        ntree      = 500,
        importance = TRUE)
))

# ── 7. Bagging (treebag) ───────────────────────────────────────────────────
models$bagging <- safe_train("Bagging (treebag)", quote(
  train(x = Xm, y = y, method = "treebag", metric = METRIC, trControl = ctrl,
        nbagg = 100)
))

# ── 8. Gradient Boosting (GBM) ─────────────────────────────────────────────
gbm_grid <- expand.grid(
  n.trees           = c(100, 300, 500),
  interaction.depth = c(1, 2, 3),
  shrinkage         = c(0.01, 0.05, 0.1),
  n.minobsinnode    = 5
)
models$gbm <- safe_train("Gradient Boosting (gbm)", quote(
  train(x = Xm, y = y, method = "gbm", metric = METRIC, trControl = ctrl,
        tuneGrid = gbm_grid, verbose = FALSE)
))

# Drop any models that failed to train
models <- Filter(Negate(is.null), models)
cat(sprintf("\nSuccessfully trained %d / 8 models\n", length(models)))
```

---

# 5. Model Comparison

## 5.1 CV performance summary table

```{r performance-table}
perf_list <- lapply(names(models), function(nm) {
  r <- models[[nm]]$results
  # Best row = row with highest ROC
  best <- r[which.max(r$ROC), , drop = FALSE]
  data.frame(
    Model       = nm,
    ROC_AUC     = round(best$ROC,  4),
    Sensitivity = round(best$Sens, 4),
    Specificity = round(best$Spec, 4),
    ROC_SD      = round(best$ROCSD, 4)
  )
})

perf_df <- do.call(rbind, perf_list) %>%
  arrange(desc(ROC_AUC))

perf_df %>%
  kable(row.names = FALSE,
        caption = "Cross-validated performance (best tuning per model)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE, color = ifelse(perf_df$ROC_AUC == max(perf_df$ROC_AUC),
                                              "green", "black"))

# Save to disk
write.csv(perf_df, file.path(RESULTS_DIR, "model_performance_summary.csv"),
          row.names = FALSE)
```

## 5.2 Resampling distribution (dotplot)

```{r resampling-dotplot, fig.height=5}
resamp <- resamples(models)
dotplot(resamp, metric = "ROC",
        main = "ROC AUC — repeated CV distribution by model")
```

## 5.3 ROC curves (cross-validated predictions)

```{r roc-curves, fig.height=6}
roc_list <- lapply(names(models), function(nm) {
  preds <- models[[nm]]$pred
  if (is.null(preds)) return(NULL)
  # Column name for positive class probability
  pos_col <- POSITIVE_CLASS
  if (!pos_col %in% colnames(preds)) return(NULL)
  roc(preds$obs, preds[[pos_col]],
      levels = rev(levels(preds$obs)),
      quiet  = TRUE)
})
names(roc_list) <- names(models)
roc_list <- Filter(Negate(is.null), roc_list)

# Build colour palette
cols <- scales::hue_pal()(length(roc_list))

# Plot
plot(roc_list[[1]], col = cols[1], lwd = 2,
     main = "Cross-validated ROC curves — all models")
for (i in seq_along(roc_list)[-1]) {
  plot(roc_list[[i]], col = cols[i], lwd = 2, add = TRUE)
}
legend("bottomright",
       legend = paste0(names(roc_list), " (AUC=",
                       round(sapply(roc_list, auc), 3), ")"),
       col    = cols, lwd = 2, cex = 0.75)
```

---

# 6. Feature Importance

```{r feature-importance, fig.height=8}
TOP_N <- 20   # number of top ASVs to display per model

extract_importance <- function(model_obj, model_name) {
  tryCatch({
    vi <- varImp(model_obj, scale = TRUE)$importance

    # varImp returns different structures depending on model
    if (ncol(vi) > 1) {
      # Classification: average across classes or take the "Overall" / first col
      if ("Overall" %in% colnames(vi)) {
        vi <- vi[, "Overall", drop = FALSE]
      } else {
        vi$Overall <- rowMeans(vi)
        vi <- vi[, "Overall", drop = FALSE]
      }
    }
    colnames(vi) <- "Importance"
    vi$Feature    <- rownames(vi)
    vi$Model      <- model_name
    vi <- vi[order(-vi$Importance), ]
    head(vi, TOP_N)
  }, error = function(e) {
    message(sprintf("  Could not extract importance for %s: %s", model_name, e$message))
    NULL
  })
}

imp_list <- mapply(extract_importance, models, names(models),
                   SIMPLIFY = FALSE)
imp_all  <- do.call(rbind, Filter(Negate(is.null), imp_list))

# Plot per model
if (!is.null(imp_all) && nrow(imp_all) > 0) {
  ggplot(imp_all, aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_col(fill = "steelblue", alpha = 0.85) +
    coord_flip() +
    facet_wrap(~ Model, scales = "free_y") +
    labs(title = sprintf("Top %d ASVs by variable importance — per model", TOP_N),
         x = NULL, y = "Scaled importance") +
    theme_bw(base_size = 9)

  write.csv(imp_all, file.path(RESULTS_DIR, "feature_importance_all_models.csv"),
            row.names = FALSE)
} else {
  cat("Feature importance could not be extracted for any model.\n")
}
```

## 6.1 Consensus importance (ASVs ranked consistently across models)

```{r consensus-importance, fig.height=5}
# For each model, rank ASVs by importance (rank 1 = most important)
# Consensus score = mean rank across models (lower = more consistently important)

consensus <- imp_all %>%
  group_by(Model) %>%
  mutate(Rank = rank(-Importance, ties.method = "average")) %>%
  ungroup() %>%
  group_by(Feature) %>%
  summarise(
    Mean_Rank       = mean(Rank),
    N_Models        = n(),
    Mean_Importance = mean(Importance),
    .groups = "drop"
  ) %>%
  filter(N_Models >= max(2, floor(length(models) / 2))) %>%  # in ≥ half the models
  arrange(Mean_Rank) %>%
  head(TOP_N)

if (nrow(consensus) > 0) {
  ggplot(consensus,
         aes(x = reorder(Feature, -Mean_Rank), y = Mean_Importance)) +
    geom_col(fill = "darkorange", alpha = 0.85) +
    coord_flip() +
    labs(title = sprintf("Consensus top ASVs (appearing in ≥ %d models)",
                         max(2, floor(length(models) / 2))),
         x = NULL, y = "Mean scaled importance") +
    theme_bw()

  write.csv(consensus, file.path(RESULTS_DIR, "consensus_feature_importance.csv"),
            row.names = FALSE)
} else {
  cat("No ASVs appeared in enough models for a consensus ranking.\n")
}
```

---

# 7. Save Results

```{r save-results}
saveRDS(models,  file.path(RESULTS_DIR, "trained_models.rds"))
saveRDS(resamp,  file.path(RESULTS_DIR, "resamples.rds"))

cat("Saved to:", RESULTS_DIR, "\n")
cat("  trained_models.rds\n")
cat("  resamples.rds\n")
cat("  model_performance_summary.csv\n")
cat("  feature_importance_all_models.csv\n")
cat("  consensus_feature_importance.csv\n")
```































######### ATTEMPT 2 #########




---
title: "ASV ML Classification: chronic_rhinitis24m_current"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Overview

This document trains a suite of ML classifiers to predict
**chronic_rhinitis24m_current** (binary: No/Yes) using ASV relative abundances
as predictors.

Key methodological choices:

- **Grouped k-fold CV by Subject_ID** — all timepoints from the same subject
  stay in the same fold, preventing data leakage across repeated measures.
- **Upsampling** within each training fold to address class imbalance (~30% Yes).
- **Near-zero variance filtering** (`nzv`) applied inside each fold.
- Primary metric: **ROC AUC**.

---

# 1. Libraries & Configuration

```{r libraries}
if (!is.null(dev.list())) dev.off()
gc()

library(tidyverse)
library(phyloseq)
library(caret)
library(glmnet)
library(randomForest)
library(pls)
library(earth)
library(gbm)
library(ipred)
library(pROC)
library(knitr)
library(kableExtra)

# ── USER CONFIGURATION ────────────────────────────────────────────────────────
PHYLOSEQ_PATH  <- "your_phyloseq.rds"   # <-- update this path
OUTCOME_COL    <- "chronic_rhinitis24m_current"
SUBJECT_ID_COL <- "Subject_ID"          # <-- update if your column name differs
POSITIVE_CLASS <- "Yes"                 # positive event label after recoding
SEED           <- 123
CV_REPEATS     <- 5
RESULTS_DIR    <- "results_asv_ml"
# ─────────────────────────────────────────────────────────────────────────────

dir.create(RESULTS_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

# 2. Load & Prepare Data

## 2.1 Load phyloseq and convert to relative abundance

```{r load-data}
ps <- readRDS(PHYLOSEQ_PATH)
cat("Phyloseq object loaded:\n")
print(ps)

ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
cat("\nConverted to relative abundance.\n")
```

## 2.2 Extract ASV feature matrix

```{r feature-matrix}
asv_mat <- as.data.frame(otu_table(ps_rel))
if (taxa_are_rows(ps_rel)) {
  asv_mat <- as.data.frame(t(asv_mat))
}

# make.names() prevents caret errors from special characters in ASV IDs
colnames(asv_mat) <- make.names(colnames(asv_mat), unique = TRUE)
cat(sprintf("ASV matrix: %d samples x %d ASVs\n", nrow(asv_mat), ncol(asv_mat)))
```

## 2.3 Extract outcome and Subject ID

```{r extract-outcome}
meta <- as.data.frame(sample_data(ps_rel))

# ── Outcome ──────────────────────────────────────────────────────────────────
if (!OUTCOME_COL %in% colnames(meta)) {
  stop(sprintf("Outcome column '%s' not found. Available columns:\n%s",
               OUTCOME_COL, paste(colnames(meta), collapse = ", ")))
}

# as.character() handles logical TRUE/FALSE, numeric 0/1, or Yes/No uniformly
outcome_raw <- as.character(meta[[OUTCOME_COL]])

# Recode to valid R variable names.
# TRUE/FALSE are reserved R keywords — caret throws an error if factor levels
# match them. Mapping to Yes/No fixes this for all input formats.
outcome_raw <- dplyr::recode(outcome_raw,
                              "TRUE"  = "Yes",
                              "FALSE" = "No",
                              "1"     = "Yes",
                              "0"     = "No")

# Explicit level order: No first, Yes second.
# twoClassSummary() treats the LAST level as the positive event.
outcome_factor <- factor(outcome_raw, levels = c("No", "Yes"))

cat("Class distribution:\n")
print(table(outcome_factor, useNA = "always"))
cat(sprintf("\nPositive class (Yes) prevalence: %.1f%%\n",
            100 * mean(outcome_factor == "Yes", na.rm = TRUE)))

# ── Subject ID ───────────────────────────────────────────────────────────────
if (!SUBJECT_ID_COL %in% colnames(meta)) {
  stop(sprintf("Subject ID column '%s' not found. Available columns:\n%s",
               SUBJECT_ID_COL, paste(colnames(meta), collapse = ", ")))
}
subject_ids <- as.character(meta[[SUBJECT_ID_COL]])
```

## 2.4 Align samples and create final X / y / groups

```{r align-samples}
shared_samples <- intersect(rownames(asv_mat), rownames(meta))
cat(sprintf("Samples in ASV matrix : %d\n", nrow(asv_mat)))
cat(sprintf("Samples in metadata   : %d\n", nrow(meta)))
cat(sprintf("Shared samples        : %d\n", length(shared_samples)))

X           <- asv_mat[shared_samples, , drop = FALSE]
y           <- outcome_factor[match(shared_samples, rownames(meta))]
subject_vec <- subject_ids[match(shared_samples, rownames(meta))]

# Remove rows with missing outcome
complete_idx <- !is.na(y)
X           <- X[complete_idx, , drop = FALSE]
y           <- droplevels(y[complete_idx])
subject_vec <- subject_vec[complete_idx]

cat(sprintf("\nFinal dataset  : %d samples x %d ASVs\n", nrow(X), ncol(X)))
cat(sprintf("Unique subjects: %d\n", length(unique(subject_vec))))
cat("Final class distribution:\n")
print(table(y))

Xm <- as.matrix(X)
n  <- nrow(X)
p  <- ncol(X)
```

---

# 3. Cross-Validation Setup

```{r train-control}
n_subjects <- length(unique(subject_vec))

# k: hold out at least 5 subjects per fold; cap at 10
k <- min(10, max(3, floor(n_subjects / 5)))
cat(sprintf("Subjects: %d  =>  %d-fold grouped CV, %d repeats\n",
            n_subjects, k, CV_REPEATS))

# ── Build grouped fold indices ────────────────────────────────────────────────
# groupKFold() ensures all rows from the same subject land in the same fold,
# so the same subject is never split across train and test.
#
# We call it CV_REPEATS times with different seeds and concatenate the results.
# Passing the full list as 'index' to trainControl with method = "cv" gives
# repeated grouped CV: caret treats each element of 'index' as one CV
# iteration, for a total of k * CV_REPEATS iterations.
set.seed(SEED)
fold_index <- unlist(
  lapply(seq_len(CV_REPEATS), function(r) {
    set.seed(SEED + r)
    groupKFold(subject_vec, k = k)
  }),
  recursive = FALSE
)

# Unique names are required by caret
names(fold_index) <- paste0(
  "Fold", formatC(rep(seq_len(k), CV_REPEATS), width = 2, flag = "0"),
  ".Rep", rep(seq_len(CV_REPEATS), each = k)
)

cat(sprintf("Total CV iterations: %d (%d folds x %d repeats)\n\n",
            length(fold_index), k, CV_REPEATS))

ctrl <- trainControl(
  method          = "cv",        # 'cv' because we supply pre-built index
  index           = fold_index,  # grouped repeated folds (no subject leakage)
  classProbs      = TRUE,        # required for ROC
  summaryFunction = twoClassSummary,
  sampling        = "up",        # upsample minority class in TRAIN folds only;
                                 # test fold class distribution is never altered
  savePredictions = "final",
  returnResamp    = "all",
  allowParallel   = TRUE,
  verboseIter     = FALSE
)

METRIC <- "ROC"
```

> **Why upsampling and not SMOTE?**
> Upsampling randomly duplicates existing minority-class rows in the training
> fold until classes are equal. SMOTE instead interpolates between real minority
> samples to create synthetic ones. For relative abundance compositional data,
> interpolation is biologically questionable — a synthetic microbiome profile
> halfway between two real ones is not a meaningful observation. Upsampling is
> the safer choice.
>
> **Why grouped folds?**
> Each subject contributes up to three timepoints (0, 4, 12 months). Standard
> CV would routinely put one timepoint in training and another in the test set
> for the same subject, effectively letting the model "see" that subject during
> training and inflating ROC estimates. Grouped folds prevent this entirely.

---

# 4. Train Models

```{r train-models, results='hold'}
set.seed(SEED)
models      <- list()
lambda_grid <- 10^seq(-4, 0, length.out = 20)

safe_train <- function(label, expr) {
  cat(sprintf("Training %s ...\n", label))
  tryCatch({
    m <- eval(expr)
    cat(sprintf("  OK  best ROC AUC: %.4f\n",
                max(m$results[[METRIC]], na.rm = TRUE)))
    m
  }, error = function(e) {
    cat(sprintf("  FAIL: %s\n", e$message))
    NULL
  })
}

# nzv removes near-zero variance ASVs within each training fold before fitting
PP <- c("nzv", "center", "scale")

# ── 1. Ridge (alpha = 0) ─────────────────────────────────────────────────────
models$ridge <- safe_train("Ridge (glmnet, alpha=0)", quote(
  train(x = Xm, y = y, method = "glmnet", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(alpha = 0, lambda = lambda_grid),
        preProcess = PP, family = "binomial")
))

# ── 2. Lasso (alpha = 1) ─────────────────────────────────────────────────────
models$lasso <- safe_train("Lasso (glmnet, alpha=1)", quote(
  train(x = Xm, y = y, method = "glmnet", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(alpha = 1, lambda = lambda_grid),
        preProcess = PP, family = "binomial")
))

# ── 3. Elastic Net ───────────────────────────────────────────────────────────
models$elastic_net <- safe_train("Elastic Net (glmnet, alpha tuned)", quote(
  train(x = Xm, y = y, method = "glmnet", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(alpha  = c(0.25, 0.5, 0.75),
                                 lambda = lambda_grid),
        preProcess = PP, family = "binomial")
))

# ── 4. PLS-DA ────────────────────────────────────────────────────────────────
max_comp <- min(15, p, n - 2)
models$plsda <- safe_train("PLS-DA", quote(
  train(x = Xm, y = y, method = "pls", metric = METRIC, trControl = ctrl,
        tuneGrid   = data.frame(ncomp = 1:max_comp),
        preProcess = PP)
))

# ── 5. MARS ──────────────────────────────────────────────────────────────────
models$mars <- safe_train("MARS (earth)", quote(
  train(x = Xm, y = y, method = "earth", metric = METRIC, trControl = ctrl,
        tuneGrid   = expand.grid(degree = 1:2, nprune = c(5, 10, 15, 20)),
        preProcess = PP,
        glm        = list(family = binomial))
))

# ── 6. Random Forest ─────────────────────────────────────────────────────────
mtry_grid <- sort(unique(c(2, floor(sqrt(p)), floor(p / 3), floor(p / 2))))
mtry_grid <- mtry_grid[mtry_grid >= 1 & mtry_grid <= p]
models$random_forest <- safe_train("Random Forest", quote(
  train(x = Xm, y = y, method = "rf", metric = METRIC, trControl = ctrl,
        tuneGrid   = data.frame(mtry = mtry_grid),
        ntree      = 500,
        importance = TRUE)
))

# ── 7. Bagging ───────────────────────────────────────────────────────────────
models$bagging <- safe_train("Bagging (treebag)", quote(
  train(x = Xm, y = y, method = "treebag", metric = METRIC, trControl = ctrl,
        nbagg = 100)
))

# ── 8. Gradient Boosting ─────────────────────────────────────────────────────
gbm_grid <- expand.grid(
  n.trees           = c(100, 300, 500),
  interaction.depth = c(1, 2, 3),
  shrinkage         = c(0.01, 0.05, 0.1),
  n.minobsinnode    = 5
)
models$gbm <- safe_train("Gradient Boosting (gbm)", quote(
  train(x = Xm, y = y, method = "gbm", metric = METRIC, trControl = ctrl,
        tuneGrid = gbm_grid, verbose = FALSE)
))

models <- Filter(Negate(is.null), models)
cat(sprintf("\nSuccessfully trained %d / 8 models\n", length(models)))
```

---

# 5. Model Comparison

## 5.1 CV performance summary table

```{r performance-table}
perf_list <- lapply(names(models), function(nm) {
  r    <- models[[nm]]$results
  best <- r[which.max(r$ROC), , drop = FALSE]
  data.frame(
    Model       = nm,
    ROC_AUC     = round(best$ROC,   4),
    Sensitivity = round(best$Sens,  4),
    Specificity = round(best$Spec,  4),
    ROC_SD      = round(best$ROCSD, 4)
  )
})

perf_df <- do.call(rbind, perf_list) %>% arrange(desc(ROC_AUC))

perf_df %>%
  kable(row.names = FALSE,
        caption = "Cross-validated performance — best tuning per model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(perf_df$ROC_AUC == max(perf_df$ROC_AUC),
                             "green", "black"))

write.csv(perf_df,
          file.path(RESULTS_DIR, "model_performance_summary.csv"),
          row.names = FALSE)
```

## 5.2 Resampling distribution

```{r resampling-dotplot, fig.height=5}
resamp <- resamples(models)
dotplot(resamp, metric = "ROC",
        main = "ROC AUC — CV distribution by model")
```

## 5.3 ROC curves (cross-validated predictions)

```{r roc-curves, fig.height=6}
roc_list <- lapply(names(models), function(nm) {
  preds <- models[[nm]]$pred
  if (is.null(preds) || !POSITIVE_CLASS %in% colnames(preds)) return(NULL)
  roc(preds$obs, preds[[POSITIVE_CLASS]],
      levels = rev(levels(preds$obs)), quiet = TRUE)
})
names(roc_list) <- names(models)
roc_list <- Filter(Negate(is.null), roc_list)

cols <- scales::hue_pal()(length(roc_list))
plot(roc_list[[1]], col = cols[1], lwd = 2,
     main = "Cross-validated ROC curves — all models")
for (i in seq_along(roc_list)[-1])
  plot(roc_list[[i]], col = cols[i], lwd = 2, add = TRUE)
legend("bottomright",
       legend = paste0(names(roc_list),
                       " (AUC=", round(sapply(roc_list, auc), 3), ")"),
       col = cols, lwd = 2, cex = 0.75)
```

---

# 6. Feature Importance

```{r feature-importance, fig.height=8}
TOP_N <- 20

extract_importance <- function(model_obj, model_name) {
  tryCatch({
    vi <- varImp(model_obj, scale = TRUE)$importance
    if (ncol(vi) > 1) {
      vi$Overall <- if ("Overall" %in% colnames(vi)) vi$Overall else rowMeans(vi)
      vi <- vi[, "Overall", drop = FALSE]
    }
    colnames(vi) <- "Importance"
    vi$Feature   <- rownames(vi)
    vi$Model     <- model_name
    head(vi[order(-vi$Importance), ], TOP_N)
  }, error = function(e) {
    message(sprintf("  Could not extract importance for %s: %s",
                    model_name, e$message))
    NULL
  })
}

imp_all <- do.call(rbind,
  Filter(Negate(is.null),
         mapply(extract_importance, models, names(models), SIMPLIFY = FALSE)))

if (!is.null(imp_all) && nrow(imp_all) > 0) {
  print(
    ggplot(imp_all, aes(x = reorder(Feature, Importance), y = Importance)) +
      geom_col(fill = "steelblue", alpha = 0.85) +
      coord_flip() +
      facet_wrap(~ Model, scales = "free_y") +
      labs(title = sprintf("Top %d ASVs by variable importance — per model", TOP_N),
           x = NULL, y = "Scaled importance") +
      theme_bw(base_size = 9)
  )
  write.csv(imp_all,
            file.path(RESULTS_DIR, "feature_importance_all_models.csv"),
            row.names = FALSE)
}
```

## 6.1 Consensus importance

ASVs are ranked within each model (rank 1 = most important), then the mean rank
is computed across models. Only ASVs appearing in at least half the trained
models are shown.

```{r consensus-importance, fig.height=5}
min_models <- max(2, floor(length(models) / 2))

consensus <- imp_all %>%
  group_by(Model) %>%
  mutate(Rank = rank(-Importance, ties.method = "average")) %>%
  ungroup() %>%
  group_by(Feature) %>%
  summarise(
    Mean_Rank       = mean(Rank),
    N_Models        = n(),
    Mean_Importance = mean(Importance),
    .groups = "drop"
  ) %>%
  filter(N_Models >= min_models) %>%
  arrange(Mean_Rank) %>%
  head(TOP_N)

if (nrow(consensus) > 0) {
  print(
    ggplot(consensus,
           aes(x = reorder(Feature, -Mean_Rank), y = Mean_Importance)) +
      geom_col(fill = "darkorange", alpha = 0.85) +
      coord_flip() +
      labs(title = sprintf("Consensus top ASVs (appearing in >= %d / %d models)",
                           min_models, length(models)),
           x = NULL, y = "Mean scaled importance") +
      theme_bw()
  )
  write.csv(consensus,
            file.path(RESULTS_DIR, "consensus_feature_importance.csv"),
            row.names = FALSE)
} else {
  cat("No ASVs appeared in enough models for a consensus ranking.\n")
}
```

---

# 7. Save Results

```{r save-results}
saveRDS(models, file.path(RESULTS_DIR, "trained_models.rds"))
saveRDS(resamp, file.path(RESULTS_DIR, "resamples.rds"))

cat("Saved to:", RESULTS_DIR, "\n")
cat("  trained_models.rds\n")
cat("  resamples.rds\n")
cat("  model_performance_summary.csv\n")
cat("  feature_importance_all_models.csv\n")
cat("  consensus_feature_importance.csv\n")
```

---

# 8. Session Info

```{r session-info}
sessionInfo()
```

